DATA 621---Business Analytics and Data Mining

### 2020-09-13 HW1/HW1_AF.Rmd
another round of editing

```{r loadData, include=FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')
```

@@

We are given a data set containing 2,276 records, each of which includes 15 statistics
and the total number of wins a team had in a given year. On average, about 50% of games 
played are won (81 games out of 162), with the best season having 146 wins and the worst 
season having 0 wins. The data is normally distributed and most years have between 49 
and 112 wins (blue lines below). The nature of the distribution means there aren't too 
many extreme seasons where wins are significantly higher or lower than usual. This serves 
as a good gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

```{r wins, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 5, fig.align = 'center'}

@@

```{r pctNull, include = FALSE}

@@

pct_null %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0)

@@

### Do the individual stats affect winning?
Intuitively, we expect that Caught stealing, Errors, Hits allowed, Homeruns allowed, 
Strikeouts by batters, and Walks allowed would all have a **negative** impact on
the total wins. In other words, as these values increase, we expect that the team 
is less likely to win.

However, when we take a closer look at the data, these relationships aren't obvious. 
In fact, only Errors and Hits allowed seem to have a negative impact on wins. Caught 
stealing and Strikeouts by batters appear to be random; this means that whether the 
stat for a particular season is high or low doesn't affect the overall number of wins.

@@

We can look at the same information for the stats that we expect to have a **positive**
effect on wins: Base hits, Doubles, Triples, Homeruns, Walks, Batters getting hit by 
pitches, Stolen bases, Double Plays, and Strikeouts by pitchers.

@@

@@

```{r fewOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 3, fig.width = 5, fig.align = 'center'}

@@

* Homeruns allowed (average = ~100/year) and Homeruns by batters (average = ~106/year) 
have a very similar mid-range distribution (50% of the data lies between ~50 and 150). 
The slight difference in average stats means that teams tend to have a higher number
of homeruns than the opposition team. 
* The only thing that stands out about strikeouts by batters (average = ~736/year) is 
how nearly perfectly normal it is. 50% of the data is between about 500 and 1000 and 
there are absolutely no outliers in the dataset! This means that there were no surprisingly
high or low seasons.

Conversely, some of the stats have a very high number of outliers, indicating that there
are some seasons with some abnormally high or low values. **Walks by batters**, **Stolen**
**bases**, **Hits allowed**, and **Errors** are examples of this.

```{r manyOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 5, fig.width = 5, fig.align = 'center'}

@@

vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~ stat, scale = 'free')

@@

>>>>

```{r loadData, include=FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(ggplot2)
library(corrplot)
library(data.table)
train_set <- read.csv('data/moneyball-training-data.csv')
eval_set <- read.csv('data/moneyball-evaluation-data.csv')
```

@@

We are given a data set of 2,276 records containing 15 seasonal statistics and the
total number of wins a team had in a given year. On average, about 50% of games 
played are won (81 games out of 162), with the best season having 146 wins and the worst 
season having 0 wins. The data is normally distributed and most years have between 49 
and 112 wins (blue lines below). The nature of the distribution means there aren't too 
many extreme seasons where wins are significantly higher or lower than usual. This serves 
as a good gut-check for our final predictions; if the predicted wins are too high or too 
low, we know something in our model is probably off.

```{r wins, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 5, fig.align = 'center'}

@@

```{r pctNull, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2, fig.width = 5, fig.align = 'center'}

@@

totalNulls <- pct_null %>%
  mutate(VARIABLE = rownames(.)) %>%
  arrange(desc(PCT_NULL)) %>%
  filter(PCT_NULL > 0) %>%
  select(VARIABLE, PCT_NULL)

ggplot(totalNulls, aes(x=reorder(VARIABLE, PCT_NULL), y=PCT_NULL, label = round(PCT_NULL, 2))) + 
  geom_text(vjust = 0.5, hjust = -0.05)+
  geom_bar(stat = "identity") +
  ggtitle("Variables with Missing Information") +
  xlab("Statistic") + ylab("Percent Missing") + 
  coord_flip() +
  expand_limits(y = 1)

@@

### Do the individual stats affect winning?

**Stats with an expected negative impact: ** Intuitively, we expect that Caught 
stealing, Errors, Hits allowed, Homeruns allowed, Strikeouts by batters, and Walks 
allowed would all have a **negative** impact on the total wins. In other words, as 
these values increase, we expect that the team is less likely to win.

@@

When we take a closer look at the data, these negative relationships aren't obvious. 
In fact, only Errors and Hits allowed seem to have a negative impact on wins. Caught 
stealing and Strikeouts by batters appear to be random; this means that whether the 
stat for a particular season is high or low doesn't affect the overall number of wins.

**Stats with an expected positive impact: ** We can look at the same information for the 
stats that we expect to have a **positive** effect on wins: Base hits, Doubles, Triples, 
Homeruns, Walks, Batters getting hit by pitches, Stolen bases, Double Plays, and Strikeouts 
by pitchers.

@@

Before using any of the statistics in a model, we need to take a closer look at the 
variation in the data. We call out of the ordinary values (exceptionally high or
low values) **outliers**. We need to take these into account in our modeling because
we want to make sure our predictions aren't skewed because of them.

@@

```{r fewOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 5, fig.align = 'center'}

@@

* **Homeruns allowed** (average = ~100/year) and **Homeruns by batters** (average = ~106/year) 
have a very similar mid-range distribution (50% of the data lies between ~50 and 150). 
The slight difference in average stats means that teams tend to have a higher number
of homeruns than the opposition team. 
* The only thing that stands out about **Strikeouts by batters** (average = ~736/year) is 
how nearly perfectly normal it is. 50% of the data is between about 500 and 1000 and 
there are absolutely no outliers in the dataset! This means that there were no surprisingly
high or low seasons.

Conversely, some of the stats have a very high number of outliers, indicating that there
are some seasons with some abnormally high or low values. **Errors**, **Hits allowed**,
**Stolen bases**, and **Walks by batters** are examples of this.

```{r manyOutliers, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 2.5, fig.width = 6, fig.align = 'center'}

@@

vals2 <- ggplot(set2 , aes(x=stat, y=value)) + 
    geom_boxplot() +
  labs(title="",x="", y = "Value") +
  facet_wrap(~ stat, scale = 'free', ncol = 4)

@@

* All of the outliers for **Errors** and **Hits allowed** are above the upper 
tail of the data set. This is further illustrated by the mean and median values
for both of these stats; in both instances, the mean per year (Errors = ~246/year
and Hits allowed = ~1779/year) are higher than the median per year (Errors = 
~159/year and Hits allowed = ~1518/year). This means that some seasons with
exceptionally high values for both of these statistics skew the dataset. 
* The maxiumum value
